A parametricity-based formalization of semi-simplicial and semi-cubical sets

### 2023-01-20 paper/paper.tex
progress

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational equality of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has a cubical flavor~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates instead of relations, in which case we obtain a form of realizability~\cite{bernardy12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication with Thorsten Altenkirch and Hugo Moeneclaey}.

@@

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \times X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions. Based on these coherence conditions, the type of $X_2$ happens to have the form $\Pi a b c d, X_1 (a, b) \times X_1 (a, c) \times X_1 (b, d) \times X_1 (c, d) \rightarrow \U$, recovering the square structure of $X_2$. In summary,

\begin{align*}
  X_0 & :                  & \U                                                       \\
  X_1 & :                  & X_0 \times X_0 \rightarrow \U                            \\
  X_2 & : \Pi a b c d, & X_1 (a, b) \times X_1 (a, c)  \times X_1 (b, d) \times X_1 (c, d) \\
 & &  \rightarrow \U                    \\
  \ldots
\end{align*}

Now, let's observe that the process of construction of the type of
$X_1$ from that of $X_0$ and from that of $X_2$ to that of $X_1$ is
similar to applying a binary parametricity translation on contexts that
is: to triplicate each declaration $x_L:X_0$ and $x_R:X_0$ into two
copies of $x$ and a line connecting these copies, that is into two
triples of declarations $x_{LL}:X_0, x_{LR}:X_0,
x_{L\kstar}:X_1(x_{LL},x_{LR})$ and $x_{RL}:X_0, x_{RL}:X_0,
x_{R\kstar}:X_1(x_{RL},x_{RR})$, and to expand $\U$ into the type of a
relation over these two copies, that is, into $x_{\kstar
  L}:X_1(x_{LL},x_{RL}) \times x_{\kstar R}:X_1(x_{LR},x_{RR})
\rightarrow \U$.

The focus of our work has been to formally define such an indexed presentation of not just semi-cubical, but more generally, $\nu$-sets.

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a : A, (B a \rightarrow C)$ and $(\Sigma a : A, B a) \rightarrow C$. Let us now illustrate how we build $\fullframe[n]$ by refining types of $X_0, X_1, \ldots, X_{n - 1}$ so that they have the form $\fullframe[n] \rightarrow \U$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Each $\fullframe[n]$ is made of $n$-layers, written $\layer[n][p]$ with $p < n$, initialized with the $\unittype$ type. Then, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$. The rest of the construction is shown in Fig. \ref{fig:intuition}

>>>>

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational equality of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has a cubical flavor~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates instead of relations, in which case, we obtain a form of realizability~\cite{bernardy12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication with Thorsten Altenkirch and Hugo Moeneclaey}.

@@

If we think of $X_1$ fibered over two copies of $X_0$ in an indexed way, we get $X_1$ of type $X_0 \rightarrow X_0 \rightarrow \U$. Continuing the process, we can think of $X_2$ fibered over four copies of $X_1$ themselves fibered over two copies of $X_0$ as a dependent type. We have a priori a dependency over eight copies of $X_0$ but these are actually four due to the coherence conditions. Based on these coherence conditions, the type of $X_2$ happens to have the form $\Pi a b c d, X_1 (a)(b) \rightarrow X_1 (a)(c) \rightarrow X_1 (b)(d) \rightarrow X_1 (c)(d) \rightarrow \U$, recovering the square structure of $X_2$. In summary,

\begin{align*}
  X_0 & :                  & \U                                                       \\
  X_1 & :                  & X_0 \rightarrow X_0 \rightarrow \U                            \\
  X_2 & : \Pi a b c d, & X_1(a)(b) \rightarrow X_1(a)(c) \rightarrow X_1 (b)(d) \\
  & & \rightarrow X_1 (c)(d) \rightarrow \U                    \\
  \ldots
\end{align*}

We observe that the process of construction of the type of $X_1$ from that of $X_0$, and from the type of $X_2$ to that of $X_1$, is similar to applying a binary parametricity translation and expecting the resulting translation to be inhabited. Indeed, parametricity interprets a type $A$ by a relation $A_\kstar$ over this type, and a term $t: A$ as an evidence that $A_\kstar(t)(t)$ is inhabited. Technically, $\U$ is interpreted as the relation ${\U}_\kstar$, which takes two types $A_L$ and $A_R$ and returns the type $A_L \rightarrow A_R \rightarrow \U$, while a dependent function type $\Pi a: A.\, B$ is interpreted as the relation $(\Pi a: A.\, B)_\kstar$ that takes two functions $f_L$ and $f_R$ of type $\Pi a: A.\, B$, and returns the $\Pi$ type taking as arguments two copies of $a: A$ together with a line between the two copies. That is,

\begin{align*}
  & (\Pi a: A.\, B)_\kstar(f_L)(f_R) \defeq \\
  & \Pi a_L: A.\, \Pi a_R: A.\, A_\kstar(a_L)(a_R) \rightarrow B_\kstar(f_L(a_L))(f_R(a_R))
\end{align*}

In particular, for $X: \U$, applying the parametricity translation is about exhibiting a relation $X_\kstar$ which inhabits ${\U}_\kstar(X)(X)$ i.e. $X \rightarrow X \rightarrow \U$. In turn, the parametricity translation of $X_\kstar: X \rightarrow X \rightarrow \U$ is about finding an inhabitant $X_{\kstar\kstar}$ of $(X \rightarrow X \rightarrow \U)\_\kstar(X_\kstar)(X_\kstar)$ i.e. of:
%. Let us equivalently write the type of $X_1$ as $\Pi x_L: X_0.\, \Pi x_R: X_0.\, \U$. To define its translation, we need to triplicate each of $x_L: X_0$ and $x_R: X_0$ into two copies and a line connecting these copies, leading to the sequences of premises $x_{LL}: X_0$, $x_{LR}: X_0$, $x_{L\kstar}: X_1(x_{LL}$, $x_{LR})$ and $x_{RL}: X_0$, $x_{RL}: X_0$, $x_{R\kstar}: X_1(x_{RL})(x_{RR})$. Then, in the context of $x_L$ and $x_R$, interpreting $X_1(x_L)(x_R): \U$ itself leads to exhibiting an inhabitant of $X_1(x_{LL})(x_{LR}) \rightarrow X_1(x_{RL})(x_{RR}) \rightarrow \U$, therefore leading to exhibiting an inhabitant of:

\begin{align*}
  \Pi x_{LL}: X.\, \Pi x_{LR}: X\, X_\kstar(x_{LL})(x_{LR}) \\
  \rightarrow \Pi x_{RL}: X.\, \Pi x_{RR}: X.\, X_\kstar(x_{RL})(x_{RR}) \\ \rightarrow X_\kstar(x_{LL}, x_{RL}) \rightarrow X_\kstar(x_{LR})(x_{RR}) \\
  \rightarrow \U
\end{align*}

which hints us at how the sequence $X_0$, $X_1$, $X_2$ can be seen as a sequence of inhabitants of the iteration $X_0 \defeq X$, $X_1 \defeq X_\kstar$, $X_2 \defeq X_{\kstar\kstar}$ of parametricity applied to an initial $X: \U$. This tells us how the previous (informal) type given to $X_2$ could be rephrased so that it comes as the instance of a general recipe characterizing the type of all $X_i$.

The focus of our work has then been to formally define such an indexed presentation of not just semi-cubical, but more generally, $\nu$-sets.

@@

We assign every $X_n$ the type $\fullframe[n] \rightarrow \U$ uniformly, applying the isomorphism between $A \rightarrow B \rightarrow C$ and $A \times B \rightarrow C$, or between $\Pi a : A, (B a \rightarrow C)$ and $(\Sigma a : A, B a) \rightarrow C$. Let us now illustrate how we build $\fullframe[n]$ by refining types of $X_0, X_1, \ldots, X_{n - 1}$ so that they have the form $\fullframe[n] \rightarrow \U$. To begin, let us set $\fullframe[0] = \unittype$, so that the type $\U$ of $X_0$ can be equivalently formulated as $\unittype \rightarrow \U$. Each $\fullframe[n]$ is made of $n$-layers, written $\layer[n][p]$ with $p < n$, initialized with the $\unittype$ type. Then, $X_1$ is made of one layer, so that it can be written as a $\Sigma$-type of an inhabitant of the $\unittype$ and $\layer[1][0]$. The rest of the construction is shown in Fig. \ref{fig:intuition}.
